---
output: 
    word_document:
        reference_docx: template.docx
---

```{r setup, message = FALSE, echo = FALSE}
devtools::load_all()
library(pls)
library(magrittr)
knitr::opts_chunk$set(echo = FALSE)
```

Word count: 350 words

`r stringr::str_to_upper("**A specific serum fatty acid (FA) composition across multiple lipid fractions underlies pathophysiological features of type 2 diabetes in the Canadian Prospective Metabolism and Islet Cell Evaluation (PROMISE) cohort**")`

**Luke W. Johnston^1^, Zhen Liu^1^, Stewart B. Harris^2^, Ravi Retnakaran^3,4^, Richard P. Bazinet^1^, Anthony J. Hanley^1,4,5^**

| ^1^ Department of Nutritional Sciences, University of Toronto, Toronto, Canada
| ^2^ Centre for Studies in Family Medicine, University of Western Ontario, London, Canada.
| ^3^ Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Canada
| ^4^ Division of Endocrinology, University of Toronto, Toronto, Canada.
| ^5^ Dalla Lana School of Public Health, University of Toronto, Toronto, Canada

**Background**: 
FA can positively or negatively influence metabolic function and could modulate
risk for diabetes. Since FA within specific lipid fractions fulfill distinct
physiological functions, a specific FA composition could be indicative of a
greater risk for diabetes. Our aim was to therefore identify specific FA
compositions across multiple lipid fractions that underlie insulin sensitivity
(IS) and beta-cell function.

**Methods**: 
Adults at risk for diabetes (n=477) had blood drawn from an OGTT. FA from
triacylglycerol (TGFA), phospholipid, cholesteryl ester, and non-esterified
(NEFA) fractions were quantified from fasting samples and the mole percent
(mol%) of total lipids was calculated. Outcome measures were the Matsuda index
(ISI) for IS and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) for
beta-cell function. Partial least squares (PLS) was used to identify underlying
clusters in the FA composition, with the FA from all lipid fractions as the
predictor variables and ISI or ISSI-2 as the response variables, for two
separate PLS models.

```{r inline}
# Models
fit_isi <- pls_model(project_data, outcome = "lISI", ncomp = 4) 
fit_issi2 <- pls_model(project_data, outcome = "lISSI2", ncomp = 4) 

# Explained Variance
# ISI:
Y_explvar_isi <- (100 * drop(pls::R2(fit_isi, estimate = "train", intercept = FALSE)$val))[2] %>% 
    round(1) %>% as.character()
X_explvar_isi <- explvar(fit_isi)[1:2] %>% 
    sum() %>% round(1) %>% as.character()
    
# ISSI-2
Y_explvar_issi2 <- (100 * drop(pls::R2(fit_issi2, estimate = "train", intercept = FALSE)$val))[2] %>% 
    round(1) %>% as.character()
X_explvar_issi2 <- explvar(fit_issi2)[1:2] %>% 
    sum() %>% round(1) %>% as.character()

# Largest contributors
lc_isi <- large_contributors(fit_isi)$fattyacid %>% 
    grep("TG", ., value = TRUE)
lc_issi2 <- large_contributors(fit_issi2)$fattyacid %>% 
    grep("TG", ., value = TRUE) 
tg_contrib <- intersect(lc_isi, lc_issi2) %>% 
    stringr::str_replace_all(" ", "") %>% 
    paste(collapse = ", ")
```

**Results**:
The first two PLS components (C1 and C2) were extracted. These components
explained `r Y_explvar_isi`% and `r Y_explvar_issi2`% of the variance in the
outcomes and explained `r X_explvar_isi`% and `r X_explvar_issi2`% of the
variance in the FA for ISI and ISSI-2, respectively. TGFA loaded heavily on C1,
while NEFA loaded heavily on C2. Only specific TGFA contributed substantially to
the variability in the outcomes (i.e. `r tg_contrib`). A higher mol% of these
specific TGFA associated with lower ISI and ISSI-2 values.

**Conclusions**:
We identified a cluster of TGFA that had a strong relationship with lower IS
and, to a lesser extent, beta-cell function. These TGFA (e.g. 16:1n-7) are
reported to associate with refined or simple carbohydrate intake and have
experimentally been shown to exhibit lipotoxic effects. Our results suggest that
only a few FA from a broad spectrum of serum FA may underlie the association
between lipids and glucose dysregulation, and that these lipids are possible
mediators between the reported associations of carbohydrate intake and diabetes
risk.

**Author information**

- Presenting author contact:
    - Name: Luke W. Johnston
    - Telephone: +16474606580
    - Email: luke.johnston@mail.utoronto.ca
- Presentation preferred: Either oral or poster
- I am a younger researcher: Post Doctorate researcher

