---
output: 
    word_document:
        reference_docx: template.docx
---

```{r setup, message = FALSE, echo = FALSE}
devtools::load_all()
library(pls)
library(dplyr)
knitr::opts_chunk$set(echo = FALSE)
```

Word count: 343 / 350 words

`r stringr::str_to_upper("**Specific clusters of fatty acid (FA) across multiple serum lipid fractions underlies pathophysiological features of type 2 diabetes in the Canadian Prospective Metabolism and Islet Cell Evaluation (PROMISE) cohort**")`

**Luke W. Johnston^1^, Zhen Liu^1^, Ravi Retnakaran^2,3^, Stewart B. Harris^5^, Richard P. Bazinet^1^, Anthony J. Hanley^1,3,5^**

| ^1^ Department of Nutritional Sciences, University of Toronto, Toronto, Canada
| ^2^ Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Canada
| ^3^ Division of Endocrinology, University of Toronto, Toronto, Canada.
| ^4^ Centre for Studies in Family Medicine, University of Western Ontario, London, Canada.
| ^5^ Dalla Lana School of Public Health, University of Toronto, Toronto, Canada

**Background**: 
Since FA within individual lipid fractions fulfill distinct
physiological functions, a specific clusters of FA could be indicative of a
greater risk for diabetes. Our aim, therefore, was to identify specific patterns in the 
composition of FA across multiple lipid fractions and their association with insulin sensitivity
(IS) and beta-cell function.

**Methods**: 
Adults at risk for diabetes (n=477) had blood drawn at fasting and during an OGTT. FA from
triacylglycerol (TGFA), phospholipid, cholesteryl ester, and non-esterified
(NEFA) fractions were quantified from fasting samples and the mole percent
(mol%) of total lipids was calculated. Outcome measures included the Matsuda index
(ISI) for IS and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) for
beta-cell function. Partial least squares (PLS) was used to identify underlying
clusters in the FA composition, with FA from all lipid fractions included as the
predictor variables and ISI or ISSI-2 as the response variables, in
separate PLS models.

```{r inline}
# Models
fit_isi <- pls_model(project_data, outcome = "lISI", ncomp = 4) 
fit_issi2 <- pls_model(project_data, outcome = "lISSI2", ncomp = 4) 

# Explained Variance
# ISI:
Y_explvar_isi <- (100 * drop(pls::R2(fit_isi, estimate = "train", intercept = FALSE)$val))[2] %>% 
    round(1) %>% as.character()
X_explvar_isi <- explvar(fit_isi)[1:2] %>% 
    sum() %>% round(1) %>% as.character()
    
# ISSI-2
Y_explvar_issi2 <- (100 * drop(pls::R2(fit_issi2, estimate = "train", intercept = FALSE)$val))[2] %>% 
    round(1) %>% as.character()
X_explvar_issi2 <- explvar(fit_issi2)[1:2] %>% 
    sum() %>% round(1) %>% as.character()

# Largest contributors
lc_isi <- large_contributors(fit_isi)$fattyacid %>% 
    grep("TG", ., value = TRUE)
lc_issi2 <- large_contributors(fit_issi2)$fattyacid %>% 
    grep("TG", ., value = TRUE) 
tg_contrib <- intersect(lc_isi, lc_issi2) %>% 
    stringr::str_replace_all(" ", "") %>% 
    paste(collapse = ", ")
estrange <- analyze_gee_pls() %>% 
    group_by(Yterms) %>% 
    mutate(estimate = round(estimate, 1)) %>% 
    summarize(estimate = paste(estimate, collapse = "-"))
```

**Results**:
The first two PLS components (C1 and C2) were extracted. These components
explained `r Y_explvar_issi2`-`r Y_explvar_isi`% of the variance in the
outcomes and explained `r X_explvar_issi2`-`r X_explvar_isi`% of the
variance in the FA. TGFA and NEFA strongly loaded negatively on C1 and C2, respectively (see Figure A).
Only specific TGFA contributed substantially to
the variability in the outcomes with C1 and C2 (i.e. `r tg_contrib`; see Figure B). A lower PLS score (higher mol% of TGFA)
associated with `r estrange$estimate[1]`% lower ISI and `r estrange$estimate[2]`% lower ISSI-2.

**Conclusions**:
We identified a cluster of TGFA that had a strong negative association with IS
and beta-cell function. These TGFA (e.g. 16:1n-7) are
reported to associate with de novo lipogenesis from simple carbohydrates and 
to exhibit lipotoxic effects. Our results suggest that
only a few FA from a broad spectrum of serum FA may underlie the association
between lipids and glucose dysregulation, and that these lipids are possible
mediators between the reported associations of carbohydrate intake and diabetes
risk.

```{r, fig.height=7, fig.width=5}
# Figure
library(patchwork)
library(ggplot2)
fig_loadings <- pls_model(project_data, "lISI", 2) %>% 
    plot_loadings() +
    ggtitle("A: Component loadings for each fatty acid for ISI")

fig_corr <- plot_corr_comps("lISI") +
    ggtitle("B: Scatterplot of correlation loadings for ISI")

fig_loadings + fig_corr + plot_layout(ncol = 1, heights = c(2, 4))
```


**Author information**

- Presenting author contact:
    - Name: Luke W. Johnston
    - Telephone: +16474606580
    - Email: luke.johnston@mail.utoronto.ca
- Presentation preferred: Either oral or poster
- I am a younger researcher: Post Doctorate researcher

